Type theory in type theory with single substitutions

### 2024-09-13 types_2024_postproc/p.tex
stuff

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the field studying type
systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When answering this
question, we aim for simplicity and ease of understanding.

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic. That is, we only consider well-formed,
  well-scoped and well-typed abstract syntax trees, there are no
  meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{alignat*}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignat*}
We use implicit arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly,
and $\app$ takes $i$, $A$ and $B$ implicitly. $f : X \cong Y : g$
denotes an isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with
$g\,(f\,x) = x$ and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$
and $\lam$ are second-order functions. We say that $\Pi$ binds a
$\Tm$-variable in its second $\Ty$-argument. Every binder is
represented by a second-order function, and as a result, we don't need
to talk about contexts and substitutions when describing the theory:
we re-use the binding features of the metatheory to describe those of
our object theory (this technique is also called higher-order abstract
syntax \cite{DBLP:conf/lics/Hofmann99} or logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
is no good notion of homomorphism between second-order models, thus
the semantics of a SOGAT is given by a first order generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts, and the usual results from universal algebra (e.g.\ a
category of models with finite limits and initial object, free and
co-free models) carry over to GATs
\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
theory of categories where there is a sort for objects and a sort for
morphisms double-indexed over objects.

@@

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort and operation of the SOGAT
gives rise to an operation in the GAT which is indexed by
contexts. The sorts moreoever come with functorial instantiation
(sometimes called substitution) operations, and all the operations are
natural with respect to instantiation. For those sorts which are bound
by an operation, we have context extension, meaning that contexts can
carry variables of that sort (in other words, the sort is given by a
locally representable presheaf). The parallel semantics of the GAT
\begin{alignat*}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignat*}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}). To define the notion
of model of a language (given by a SOGAT) is to take its parallel
substitution semantics, which is a GAT, and that induces a notion of
model. The syntax of the language is the initial model, which exists
for every GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an
intrinsic definition of syntax (only well-typed terms), which is
quotiented by the equations in the SOGAT, which describe conversion
for the language.

The single substitution semantics is an alternative GAT xxx

conversion to GAT, algebra, models

The traditional way of defining a programming language formally
\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) types are part of the definition and terms are indexed
by them, thus only allowing ASTs to be built in well-typed ways; (iii)
terms are quotiented by conversion rules, making convertible terms
equal. Compared to the traditional extrinsic approach, the intrinsic approach provides
shorter, less technical definitions, and there
is no mismatch between syntax and semantics. Type theory is a
generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}), the notion of model is direct,
and the syntax is simply the initial model.

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
B-systems, the listed notions all define a parallel substitution
calculus, where contexts and substitutions form a category. Like
B-systems, in this paper we define a single substitution calculus
(SSC). Single substitutions don't compose, so they do not form a
category. Our definition is simpler than B-systems, but it is closely
related to it. It is a minimalistic presentation where every operation
and equation is well-motivated (see Section \ref{sec:ssc}). In
addition to its pedagogical value due to its simplicity, in the future
it might give rise to new models or contribute to a possible coherent
syntax for type theory, see Section \ref{sec:conclusion}.

CwFs, C-systems, B-systems and our SSC only provide the basic
substitution calculus part of the theory, and individual type formers
(such as function space or product types) have to be specified
separately. There is a systematic way to do this: non-substructural
programming languages can be specified as second-order generalised
algebraic theories (SOGATs \cite{DBLP:conf/fscd/KaposiX24}), which are
GATs allowing second-order operations. For example, type theory with
$\Pi$ and a family of

Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} showed that any SOGAT
has two kinds of GAT semantics, one based on a parallel substitution
calculus, and another based on a single substitution calculus. The
parallel substitution GAT semantics of the SOGAT containing only the
sorts of types and terms ($\Ty : \Set, \Tm : \Ty\ra\Set$) is simply
the GAT of CwFs. The single substitution semantics of the same SOGAT
is the SSC in this paper.

SOGAT stuff

Our SSC is in some sense too minimalistic: for example, consider a
term $b$ which depends on a context $\Gamma$ and an extra variable of
type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so
that we add an extra $A$ variable \emph{before} the last one obtaining
$
b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])
$
where $\p$ is
weakening at the end of the context, and $\p^+$ is weakening just
before the last variable in the context. Then, we instantiate the last
$A[\p]$ variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ denotes the last
variable in the context (a.k.a\ De Bruijn index $0$).
We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
an arbitrary model, but this equation holds in the syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In fact, we can show
that the syntax xxx

In fact, all the
operations and equations of the parallel substitution calculus (CwF)
are admissible in the syntax of SSC, moreover the set of syntactic SSC
terms are isomorphic to the set of CwF terms. In contrast with
B-systems which are equivalent to CwFs, our SSCs do not form
CwFs. Every CwF-based model is a model of our SSC, but not the other
way: there are more SSC-models than CwF-models. The situation is
analogous to the relationship of extensional combinatory calculus and
lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic
terms form isomorphic sets, but there are more combinatory models than
lambda calculus models.

We define the parallel substitution calculus operations by induction
on the SSC-syntax. One of our constructors is the instantiation
operation of single substitutions: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on terms, we always have to provide a
method for these constructors, but this can be difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal forms. Just as $\beta$-normal
forms don't distinguish between $\beta$-equal terms, $\alpha$-normal
forms don't include explicit substitutions, all the variables are
instantiated. $\alpha$-normal forms are not usual normal forms in the
sense that they are still quotiented by the computation/uniqueness
rules like $\beta$/$\eta$ for function space.

In this paper, we define the SSC and extend it with
$\Pi$ types and a hierarchy of Coquand-universes
\cite{coquandUniverse}. The rules for type formers are the same as for
CwFs, the difference is only in the substitution calculus part. More
generally, every second-order generalised algebraic theory (SOGAT) has
a parallel and a single substition-based first order presentation
\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
the parallel and the single substitution calculi are isomorphic for
the following SOGAT

We conjecture that the methods of in this paper can be used to show
that for any SOGAT, the two first-order presentations give isomorphic
syntaxes.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus (SSC).
\item The $\alpha$-normalisation technique which is useful for working
  with explicit substitution calculi.
\item The proof of equivalence of the SSC-syntax with the CwF-based
  syntax.
\end{itemize}

@@

@@

\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}

@@

\subsection{Formalisation}

@@

@@

Contents of the TYPES abstract, but with more details.

>>>>

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the field studying type
systems (e.g.\ \cite{DBLP:books/daglib/0005958}).

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignedat}
\end{equation}
We use implicit arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly,
and $\app$ takes $i$, $A$ and $B$ implicitly. $f : X \cong Y : g$
denotes an isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with
$g\,(f\,x) = x$ and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$
and $\lam$ are second-order functions. We say that $\Pi$ binds a
$\Tm$-variable of type $A$ in its second $\Ty$-argument. Every binder
is represented by a second-order function, and as a result, we don't
need to talk about contexts and substitutions when describing the
theory: we re-use the binding features of the metatheory to describe
those of our object theory (this technique is also known as
higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99} or
logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
is no good notion of homomorphism between second-order models, thus
the semantics of a SOGAT is given by a first order generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts. The usual results from universal algebra (e.g.\ models
form a category with finite limits and initial object, existence of
free and co-free models) carry over to GATs
\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
theory of categories where there is a sort for objects and a sort for
morphisms double-indexed over objects.

@@

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort/operation of the SOGAT gives
rise to a sort/operation in the GAT which is now indexed by
contexts. The sorts moreoever come with instantiation (sometimes
called substitution) operations which are functorial, and all the
operations are natural with respect to instantiation. For those sorts
which are bound by an operation in the SOGAT description, we have
context extension, meaning that contexts can carry variables of that
sort (in other words, the sort is given by a locally representable
presheaf). The parallel semantics of the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignedat}
\end{equation}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}). We define the notion
of model of a language (a SOGAT) by its parallel substitution
semantics, which is a GAT, and that induces a notion of model. The
syntax of the language is the initial model, which exists for every
GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
syntax (only well-typed terms), quotiented by the equations in the
SOGAT, which describe conversion for the language.

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:ssc}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models, our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. Because our
single substitution GAT does not involve a category, it might be
easier to adapt it in a higher dimensional setting, see Section
\ref{sec:conclusion}.

The single substitution semantics of (\ref{eq:tytm}) is in some sense
too minimalistic: for example, consider a term $b$ which depends on a
context $\Gamma$ and an extra variable of type $A$, that is, $b :
\Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so that we add an extra $A$
variable \emph{before} the last one obtaining $ b[\p^+] :
\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p]) $ where $\p$ is weakening at
the end of the context, and $\p^+$ is weakening just before the last
variable in the context. Then, we instantiate the last $A[\p]$
variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$
denotes the last variable in the context (a.k.a\ De Bruijn index $0$).
We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary
model, but this equation holds in the syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In fact, we show that
the single substitution syntax of (\ref{eq:tt}) is equivalent to the
parallel substitution syntax of (\ref{eq:tt}), and the techniques that
we use here are not specific to the SOGAT (\ref{eq:tt}), we expect
that they work for any SOGAT. Equivalence of the syntaxes means that
the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
general SOGAT, there are more models of the correponding single
substitution GAT than the parallel substitution GAT. Every parallel
model is also a single model, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
substitution models of (\ref{eq:tt}) than parallel substitution
models. The situation is similar to that of extensional combinatory
calculus and lambda calculus, where the syntaxes are equivalent, and
every lambda model is a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship of monoids over a set $X$ and nil/cons algebras over $X$,
where every monoid is a nil/cons-algebra, but not the other way; it is
well-known that free monoids over $X$ (syntax for monoid over $X$) are
the same as lists of $X$ (the syntax for nil/cons algebras over $X$).

When showing that the parallel and single substitution syntaxes are
equivalent, we define the parallel operations by induction on the
single substitution syntax. One of the operations in the single
substitution syntax is the instantiation operation: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on types/terms, we always have to provide
a method for these operations, but this can be difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal types and terms. Just as
$\beta$-normal forms don't distinguish between $\beta$-equal terms,
$\alpha$-normal forms don't include explicit substitutions, all the
variables are instantiated. $\alpha$-normal forms are not usual normal
forms in the sense that they are still quotiented by the
computation/uniqueness rules like $\beta$/$\eta$ for function
space. To derive the new induction principle for the syntax, we first
prove $\alpha$-normalisation.

if we have U,Pi, we can minimise the syntax

if we have U,Pi,Sigma, then SSC semantics of (1)+Sigma -> PSC semantics of (1)+Sigma.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item Minimisation.
\item For type theory with $\Pi$, $\U$, $\top$ and $\Sigma$, deriving
  a parallel model from a single substitution model.
\end{itemize}

@@

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
B-systems, the listed notions all define a parallel substitution
calculus, where contexts and substitutions form a category.

@@

@@

\subsection{Metatheory and formalisation}

@@

Prop.

@@

Contents of the TYPES abstract, but with more details.
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
& \Sub && : \Con\ra\Con\ra\Set \\
& \blank[\blank] && : \Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta \\
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
& \Pi && : (A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
& \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
& \q[^+] && : \q[\gamma^+] = \q \\
& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
& \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
\end{alignat*}
